The variability of Betelgeuse explained by surface convection

### 2024-03-07 main.tex
modified:   art75.bib 	modified:   main.tex

Betelgeuse is a prototypical red supergiant (RSG) known to be a semi-regular variable. Several periods can be found in the 
literature usually clustered around the so-called long secondary period (LSP) of 2000 days and other more or less well defined 
periods around 400 and 200 days. Such periods are often linked to radial pulsation modes,   and have been used to model 
the characteristics of the resonant cavity, hence the size of the star and eventually its evolutionary stage \cite{XXX}.

At the end of 2019, Betelgeuse reached a historical minimum in its luminosity, called the great dimming \citep{guinan_fall_2020}. 
From interferometric data, we know that this event was caused by a cloud of dust close to the line of sight \citep{montarges_dusty_2021}. 
Interferometric images show a huge drop in luminosity in the southern hemisphere of Betelgeuse, leading to this dimming. In this event it was clear that a change in brightness was not due to any kind of pulsation but to a change on the 
brightness distribution over the stellar disk due to the presence of large quantities of dust. While this may be seen as 
a singular event, it brings up the question whether all changes in brightness cannot be due to similar changes in the brightness 
distribution over the disk, unrelated to any pulsation phenomenon.

Betelgeuse is known to present large convective cells with life times in the order of one to two years and measurable changes 
in the span of one week. Bright convection cells near disk center will increase the integrated brightness of the star when 
compared with other situations where such cells are found near the edges. So even without calling for the formation of 
dust or other changes in opacity, one may argue that the simple evolution of convective patterns of the star may create a variability. Such 
variability could be expected to be random, but will present quasi-periodicities related to the typical time scales of the 
convective patterns.

Since the end of the Great Dimming event, Betelgeuse has continue its random variation of brightness. 
Interestingly, 
it has been shown by \cite{jadlovsky_analysis_2023} and \cite{dupree_great_2022} that the periodicity of Betelgeuse has changed after the dimming. Using the light curves from AAVSO, 
the authors showed that before the dimming, the dominant period of Betelgeuse was around 400 days. However, after the dimming, this period has changed and 
is now shorter, around 200 days. It means that, after the dimming, there was a change in the behaviour of the atmosphere of Betelgeuse: the period is now
shorter than before the dimming. Our analysis is based on this change of periodicity. 
Faced with the dramatic change in stellar parameters and age that such change in a pulsation period would 
require, our interest in an alternative explanation of the variability
in terms  of convective patterns arose. In order to address this question we shall seek for such typical periods of 1000, 400 and 200 days in  
observational proxies related to the convective activite but unrelated to pulsations and which cannot be directly linked to brightness?

Linear polarization in the atomic lines of the spectra of Betelgeuse, discovered by \cite{auriere_discovery_2016}, has been interpreted as the joint action of 
two mechanisms. First the depolarization of the continuum by atoms which absorb linearly polarized light and re-emit unpolarized light, the 
continuum photons being polarized by Rayleigh scattering. This depolarization produces signals with an azimuth symmetry over the disk that 
would  cancel out the net linear polarization on a homogeneous disk. Thus it must be combined with an inhomogenous disk to produce the net linear polarization 
signal observed. This interpretation suggested the possibility of mapping those brightness inhomogeneities. This has been done by \cite{lopez_ariste_convective_2018} 
even producing 3-dimensional images of the atmosphere of Betelgeuse \citep{lopez_ariste_three-dimensional_2022} when taking advantage of the different height of formation of 
different lines in the spectrum of Betelgeuse. The images produced compare very well with cotemporaneous images made with interferometric 
techniques and show clear convective patterns, akin to solar granulation.

@@

Betelgeuse has been observed for the last 13 years with Narval and Neo-Narval at the Telescope Bernard Lyot. These two instruments measure the polarization 
over the visible spectra of Betelgeuese (390-1000 nm) with high spectral resolution (R=65000) and high polarimetric sensitivity. Despite 
this high sensitivity, the signal-to-noise ratios are not sufficient to measure the weak polarization signals in individual atomic lines of the 
spectrum. These amplitudes are today known to be of the order of $10^{-4}$ times the continuum intensity, and only excepcionnaly do they 
reach amplitudes of $10^{-3}$ the continuum. In these exceptional observations, enough photons can be accumulated per spectral bin to 
see the linear polarization signal above noise \citep{auriere_discovery_2016}. Most commonly, the amplitudes of linear polarization are below noise levels. In those 
occassions we have to add up the signals of thousands of lines to reduce the noise and increase the signal-to-noise ratios. This line addition 
is done through a technique called Least-Squares Deconvolution (LSD) \citep{donati_spectropolarimetric_1997} and assumes that the 
linear polarization signal is similar 
in all spectral lines up to scale factors in the sampling and amplitude. This technique has been successfully used in the past to measure 
magnetic field distributions over stellar surfaces and it is now used to produce images of the brightness distributions in the photosphere 
of Betelgeuse and other red supergiants as cited in the Introduction.

In this work we shall look into the signals produced through these line-addition techniques, and which produce a pseudo-spectral line in intensity 
and linear polarization which does not belong to any atomic species in particular but to an average of all of them present and emitting in the 
photosphere of Betelgeuse. After such procedures of line addition, these signals carry the coherent signals present in the photospheric lines, but the particularities of this or 
that spectral line are erased. All these data has been presented before by \cite{auriere_discovery_2016}, \cite{mathias_evolution_2018} and 
\cite{lopez_ariste_three-dimensional_2022}, which also give details on observing times and conditions as well as more detailed description 
on the data reduction \footnote[1]{Beyond a 2-year proprietary embargo, and up to technical issues, all these data is available through 
PolarBase (http://polarbase.irap.omp.eu/).}

@@

Trying to find periods in an astrophysical context carries the hurdle of unevenly spaced observed data. To overcome this issue, 
we used the Lomb-Scargle (hereafter LS) periodogram \citep{lomb_least-squares_1976,scargle_studies_1982}. For a given set of data, it fits sinusoid functions 
using least-squares at each frequency between the first and the last observation. The better the fit, the higher the power attributed to the frequency. 
We apply this technique to the  LSD profiles of Stokes I, U and Q of Betelgeuse observed at the TBL during 
that span of 13 years.

We first computed the LS periodogram at each wavelength of the LSD profile, between -25 km/s and 80 km/s to span the entire 
signal present in the LSD profile. \cite{lopez_ariste_convective_2018} found the most blueshifted signal towards -20 km/s, 
and interpreted it as  the maximum velocity of the rising plasma, seen from the Earth. The most redshifted signal was found at 40 km/s. 
This velocity is interpreted
as the rest velocity of the star in our reference frame. Seldom, signals at velocities larger than 40 km/s can be found 
and can be interpreted as either dark and cold plasma falling back to Betelgeuse, 
or to rising plumes of plasma located in the hidden face of Betelgeuse,
and rising so high that they appear above the limb.

@@

Figures \ref{LS intensity},\ref{LS Q},\ref{LS U} and \ref{LS linear polarization} show the Lomb-Scargle periodogram in normalised power of the LSD profile of respectively 
the intensity, stokes Q, stokes U and the linear polarization. 
%The abscissa is the frequency (in $\mathrm{days^{-1}}$), while the ordinate is the heliocentric radial velocity in km/s. 
For each figure, the upper panel is the LS periodogram with data taken before the great dimming of Betelgeuse, that is from 2013 to the beginning of 2019. 
The lower panel is the LS periodogram for the whole period before and after the dimming: from 2013 to 2023. The white dashed lines mark the different 
periods found in the literature: the LSP at 2000 days, the fundamental pressure mode at 400 days and its first overtone at 200 days.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle Intensity.png}
    \caption{Lomb-Scargle periodogram of the LSD profile of intensity. 
    The upper pannel is the LS periodogram for data before the great dimming while the lower panel is the periodogram from the all dataset.
    The ordinate refers to the velocity in the heliocentric radial velocity in km/s. The abscissa is the period in $\mathrm{days^{-1}}$. 
    The three white dashed lines represent respectively (from left to right) the LSP at 2000 days, the 400 days period and its first overtone at 200 days. 
    The Lomb-Scargle power refers to the quality of the fit from the LS periodogram. }
    \label{LS intensity}
\end{figure}

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle linear polarization.png}
    \caption{Same as figure \ref{LS intensity} for the linear polarization: $\sqrt{\mathrm{Q^2}+\mathrm{U^2}}$.}
    \label{LS linear polarization}
\end{figure}

@@

%\subsection{Intensity}
Figure \ref{LS intensity} shows the LS periodogram of the LSD of the intensity profile. Interestingly, we recover the LSP at 2000 days, 
but located at a HRV of $\sim$ 20 km/s rather than over the whole profile. This periodicity is still present after the  dimming, but  it is located at a HRV between 0 and 10 km/s. 
There is also a weaker signal at the fundamental mode, around 400 days before the dimming. But this signal completely disappears after the dimming. From these observations
we find  hard to attribute 
the observed LS power to a global period of Betelgeuse.
%, however it might be a hint that this periodicity has changed after the dimming. 
Our results 
in the intensity profile are consistent with the one from \cite{mathias_evolution_2018} who used 2D fourier analysis and found that the dominant period was the LSP.

@@

Figure \ref{LS Q} and \ref{LS U} show the LS periodogram of both stokes Q and U. It is clear that the LS periodogram is completely different 
from the intensity one. A strong signal is present in fig. \ref{LS Q} around 200 days. Once again, this signal is weaker after the dimming.
Concerning stokes U in fig. \ref{LS U}, it appears that we only recover the LSP, which once again, is weaker after the great dimming. 
Those periodograms rise a lot of questions: why is the 200 days period only present in stokes Q? Why not in stokes U or intensity? 
We will discuss about those questions in the next section. It also appears that stokes Q is not sensitive to the LSP, or to be more precise, 
it is much more sensitive to the 200 days period. From this, we can conclude that stokes Q evolves on a time scale much more closer to the first overtone 
mode than the LSP or even the fundamental mode. Since Stokes Q is sensitive  to different periods, we must conclude that it is due to other mechanisms
than the one behind the periodicitys observed in intensity. Linear polarization in Betelgeuse is interpreted to be a proxy of  convective cells,
so we can speculate that the signal present in the periodogram 
may be related to  the lifetime of convective cells. Such hypothesis is comforted by what we see in the 
LSD profiles of  Betelgeuse  whose shape is overhauled  in typical time scales of  a few months,  
Such time scales observed in the Stokes profiles and attributed to the convective time scales 
appear to explain the strong signal present at 200 days in the LS periodogram of Stokes Q.

@@

Figure \ref{LS linear polarization} shows the LS periodogram of the total linear polarization of Betelgeuse: $\sqrt{\mathrm{Q^2+U^2}}$. 
Interestingly, the LS periodogram shows a strong power at 400 days and 2000 days. This signal is much more weaker after the dimming.
While the LSP is already present in the LS of intensity, here we recover the 400 days periodicity, often invoked in the literature. 
Nonetheless, the power of this period peaks at a HRV of around 40 km/s. We recall that this velocity is the rest velocity of the star. 
Every signal beyond this velocity is either due to  plasma sinking in Betelgeuse, or to plumes of plasma rising beyond and above the limb of the star. 
Such plumes from the back hemisphere haven been often seen in other red supergiants like $\mu$ Cep \citep{lopez_ariste_height_2023}. 
This suggests that both signals at these period of 400 days found in intensity and total linear polarisation 
could well be due to the periodic appearance of plumes in the back himisphere rising high enough to appear above
the visible limb and generating a supplementary source of brightness. Once again we find a plausible explanation 
of these periods unrelated to pulsations, but rather to the convective dynamics of the star.

@@

Using linear polarization, \cite{lopez_ariste_convective_2018} have been able to reconstruct images of Betelgeuse that have been sucessfully compared 
to inteferometric images of \cite{montarges_close_2016}. The images are produced by finding the brightness distribution that better fits the observed linear 
polarization LSD profile using a Marquardt-Levemberg minimisation.

Betelgeuse is observed in average every other weeks by the TBL, and hence
we are able to follow its surface activity through polarimetry imaging. This has  allowed in the past the estimatation of  the size and the
 life time of the convective cells
at the surface. From these images, we can compute a photo-center, which is sensitive to the size and the number of convective cells. 
An homogeneous star will have a photo-center coinciding with barycenter of the star, whereas a star with one or two huge convective cells will have a
photo-center displacement more important, up to a few percent of the stellar radius in the case of RSG \citep{chiavassa_probing_2022}.

@@

We computed the LS periodogram of the displacement of the photo-center. But, before going any further, we have to warn
 about the interpretation of such images. 
Linear polarization suffers from a $180 ^\circ$ degree ambiguity already mentioned in \cite{auriere_discovery_2016}. For that reason, our images 
can be 
rotated by $180^\circ$ degree, and the brightness distribution will still fit the observed LSD polarization profiles. 
If a particular observation were to be treated independently of previous observations, the solution found by 
the algorithm could be any of the ambigous solutions possible. To obtain a continuity 
between the series images, 
for a given day we use the brightness distribution of the previous day as initial point 
of the  fitting iteration; and for the first image, we start with a 
random brightness distribution as the beginning of the fit. 
Even though we are able to produce a series of images  consistent with each other, it is important 
to keep in mind that the photo-center displacement computed from the series 
will be affected by the choice of the first image. 
To overcome this issue, we decided to compute
the photo-center displacement from 100 different initial parameters and their subsequent series 
of images in an effort to recover from the ensemble
properties that are independent of the choice of the first image.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle Photo-center.png}
    \caption{Lomb-Scargle periodogram of the 100 photo-center displacement of Betelgeuse. Each blue curve corresponds to a LS periodogram of one photo-center
     displacement. The black line is the average of the 100 LS periodogram. The black dashed lines represent the 2000, 400 and 200 days period respectively 
     (from left to right). The upper panel is the LS from data before the dimming. The lower panel is the LS from the all dataset, 
     including data after the great dimming. }
    \label{LS photocenter}
\end{figure}

Figure \ref{LS photocenter} shows the LS periodogram of the photo-center displacement  of each one of those 100 series (blue lines) and the average LS periodogram (black line). 
As before, the upper panel shows the LS from images before the great dimming of Betelgeuse only, whereas the lower panel is the LS for the whole set of 
data,  including images after the dimming. The black dashed lines mark the 2000, 400 and 
200 days periods respectively. First, we find the the LS power is high around the LSP, though
uncomfortably close to the auto-correlation peak.
We also notice a decrease in the Lomb-Scargle power from  before to after the dimming, in 
concurrence to what was seen in the LS periodograms of the LSD profiles. Before it, there are two smalls bumps around the 400 days period. 
But those bumps almost completely disappear after the dimming. Even though it is 
hard to identify the 400 days period from those bumps alone,
we still find that there is a difference before and after the dimming. 
And this is concluded  from images that exclusively involve surface convection.
The LS of the photo-center displacement points towards a variability explained by convection only. 
And this is consistent with what we obtained from linear polarization.
%Both the photo-center displacement and the LS of the linear polarization point toward a variability explained by surface convection.

@@

We have recovered the light curve of Betelgeuse in the visible from the AAVSO database for the last 13 years (Figure \ref{light curve Betelgeuse}).
Before the great dimming, the magnitude of Betelgeuse evolves on a timescale of a year, whereas after the dimming, the variability of Betelgeuse has changed. 
It is clear from fig. \ref{light curve Betelgeuse} that the 
variability of Betelgeuse is evolving on a timescale shorter than one year. This qualitative change has been pointed out before.
In the literature, it is reported that before the great dimming the main period of Betelgeuse was of 400 days \citep{kiss_variility_2006}. 
This period is often associated to the fundamental pressure mode. After the dimming, however this period appears to have disappeared 
and only the period at 200 days,  associated to the first overtone of the fundamental mode, is visible since \citep{XXX}.
No explanations based upon pulsations have been ofered concerning this change of variability after or perhaps because of the dimming. But this 
is a behaviour that fully concurs with what we find in the polarization data, and which cannot be related to any stellar pulsation but rather 
convective dynamics. Beyond the actual values of the periods found in a Lomb-Scargle periodogram, we also see a correspondence between polarization 
and the light curve in these changes in the observed periods before and after the great dimming. Such concurrence points to an explanation 
in terms of convective dynamics also for the variability of the light curve.

It may be argued that the periodograms presented in the previous sections and, in particular, those  of the photo-center displacements, 
present low amplitude peaks at the referred main periods. One may wonder whether we are over-interpreting mere fluctuations of the periodograms, 
while the peaks seen in the literature from analysis of the light curve are unmistakable. It is therefore worth to take advantage of the 
availability of the AAVSO data to perform a Lomb-Scargle periodogram of the light curve in the same period covered by our polarization data and,
furthermore, taking into consideration data around the dates for which TBL data is available. Periodograms  with such constraints are 
shown in Fig.\ref{LSAAVSO}. Even when the full dataset for the period is taken into account, the periods present small amplitudes, and are 
fully and favourably comparable with the periodograms presented above. This is even more dramatic when the periodogram is made from data at the exclusive 
dates at which TBL data is available. It stands out that the low amplitude peaks are rather due to both the sparse cadence and short span of 
time of the TBL data series rather than to a physical absence. Actually, the present tests puts more emphasis on the importance of the 
strong peaks seen at selected wavelengths in the LSD profiles, strong when compared to the peaks at similar periods arising from the light curve.

@@

The interpretaion of the linear polarization profiles is made in terms of convective structures in the atmosphere of Betelgeuse. We suggest 
that it is in these convective structures and their temporal evolution that we will find the true reason for the observed periodicities both 
in the light curve and in the spectropolarimetric observations. This is the main conclusion of this work: the observed periods are related 
to convective dynamics and not to radial pulsations of the star.

>>>>

Betelgeuse is a prototypical red supergiant (RSG) known to be a semi-regular variable. Several periods can be found in the 
literature usually clustered around the so-called long secondary period (LSP) of 2000 days and other more or less well defined 
periods around 400 and 200 days. Such periods are often linked to radial pulsation modes,   and have been used to model 
the characteristics of the resonant cavity, hence the size of the star and eventually its evolutionary stage \citep{saio_evolutionary_2023}.

At the end of 2019, Betelgeuse reached a historical minimum in its luminosity, called the great dimming \citep{guinan_fall_2020}. 
From interferometric data, we speculate that this event was caused by a cloud of dust close to the line of sight \citep{montarges_dusty_2021}. 
Interferometric images show a huge drop in luminosity in the southern hemisphere of Betelgeuse, leading to this dimming. In this event it was clear that a change in brightness was not due to any kind of pulsation but to a change on the 
brightness distribution over the stellar disk due to the presence of large quantities of dust. While this may be seen as 
a singular event, it brings up the question whether all changes in brightness cannot be due to similar changes in the brightness 
distribution over the disk, unrelated to any pulsation phenomenon.

Betelgeuse is known to present large convective cells with lifetimes in the order of one to two years and measurable changes 
in the span of one week. Bright convection cells near disk center will increase the integrated brightness of the star when 
compared with other situations where such cells are found near the edges. So even without calling for the formation of 
dust or other changes in opacity, one may argue that the simple evolution of convective patterns of the star may create a variability. Such 
variability could be expected to be random, but will present quasi-periodicities related to the typical time scales of the 
convective patterns.

Since the end of the Great Dimming event, Betelgeuse has continue its random variation of brightness. 
Interestingly, 
it has been shown by \cite{jadlovsky_analysis_2023} and \cite{dupree_great_2022} that the periodicity of Betelgeuse has changed after the dimming. Using the light curves from AAVSO, 
the authors showed that before the dimming, the dominant period of Betelgeuse was the 400 days period. However, after the dimming, this period has changed and 
is now shorter, around 200 days. Therefore, after the dimming, there was a change in the behaviour of the atmosphere of Betelgeuse: the time scales on which Betelgeuse evolves are now shorter. Our analysis is based on this change of periodicity. 
Faced with the dramatic change in stellar parameters and age that such change in a pulsation period would 
require, our interest in an alternative explanation of the variability
in terms  of convective patterns arose. In order to address this question we shall seek for such typical periods of 2000, 400 and 200 days in  
observational proxies related to the convective activity but unrelated to pulsations and which cannot be directly linked to brightness.

Linear polarization in the atomic lines of the spectra of Betelgeuse, discovered by \cite{auriere_discovery_2016}, has been interpreted as the joint action of 
two mechanisms. First the depolarization of the continuum by atoms which absorb linearly polarized light and re-emit unpolarized light, the 
continuum photons being polarized by Rayleigh scattering. This depolarization produces signals with an azimuth symmetry over the disk that 
would  cancel out the net linear polarization on a homogeneous disk. Thus it must be combined with an inhomogenous disk to produce the net linear polarization 
signal observed. This interpretation suggested the possibility of mapping those brightness inhomogeneities. This has been done by \cite{lopez_ariste_convective_2018} 
even producing 3-dimensional images of the atmosphere of Betelgeuse \citep{lopez_ariste_three-dimensional_2022} when taking advantage of the different height of formation of 
different lines in the spectrum of Betelgeuse. The images produced compare very well with contemporaneous images made with interferometric 
techniques \citep{montarges_close_2016} and show clear convective patterns, akin to solar granulation.

@@

Betelgeuse has been observed for the last 13 years with Narval and Neo-Narval at the Telescope Bernard Lyot (TBL). These two instruments measure the polarization 
over the visible spectra of Betelgeuese (390-1000 nm) with high spectral resolution (R=65000) and high polarimetric sensitivity. Despite 
this high sensitivity, the signal-to-noise ratios are not sufficient to measure the weak polarization signals in individual atomic lines of the 
spectrum. These amplitudes are today known to be of the order of $10^{-4}$ times the continuum intensity, and only exceptionally do they 
reach amplitudes of $10^{-3}$ the continuum. In these exceptional observations, enough photons can be accumulated per spectral bin to 
see the linear polarization signal above noise \citep{auriere_discovery_2016}. Most commonly, the amplitudes of linear polarization are below noise levels. In those 
occasions we have to add up the signals of thousands of lines to reduce the noise and increase the signal-to-noise ratios. This line addition 
is done through a technique called Least-Squares Deconvolution (LSD) \citep{donati_spectropolarimetric_1997} and assumes that the 
linear polarization signal is similar 
in all spectral lines up to scale factors in the sampling and amplitude. This technique has been successfully used in the past to measure 
magnetic field distributions over stellar surfaces and it is now used to produce images of the brightness distributions in the photosphere 
of Betelgeuse and other red supergiants as cited in the Introduction.

In this work we shall look into the signals produced through these line-addition techniques, and which produce a pseudo-spectral line in intensity 
and linear polarization which does not belong to any atomic species in particular but to an average of all of them present and emitting in the 
photosphere of Betelgeuse. After such procedures of line addition, these signals carry the coherent signals present in the photospheric lines, but the particularities of this or 
that spectral line are erased. All these data has been presented before by \cite{auriere_discovery_2016}, \cite{mathias_evolution_2018} and 
\cite{lopez_ariste_three-dimensional_2022}, which also give details on observing times and conditions as well as more detailed description 
on the data reduction. \footnote[1]{Beyond a 2-year proprietary embargo, and up to technical issues, all these data is available through 
PolarBase (http://polarbase.irap.omp.eu/).}

@@

Trying to find periods in an astrophysical context carries the hurdle of unevenly spaced observed data. To overcome this issue, 
we used the Lomb-Scargle periodogram \citep{lomb_least-squares_1976,scargle_studies_1982}. For a given set of data, it fits sinusoidal 
using least-squares at each frequency between the first and the last observation. The better the fit, the higher the power attributed to the frequency. 
We applied this technique to the  LSD profiles of Stokes I, U and Q of Betelgeuse observed at the TBL during 
that span of 13 years.

We first computed the Lomb-Scargle periodogram at each wavelength of the LSD profile, between -25 km/s and 80 km/s to span the entire 
signal present in the LSD profile. \cite{lopez_ariste_convective_2018} found the most blueshifted signal towards -20 km/s, 
and interpreted it as  the maximum velocity of the rising plasma, seen from the Earth. The most redshifted signal was found at 40 km/s. 
This velocity is interpreted
as the rest velocity of the star in our reference frame. Seldom, signals at velocities larger than 40 km/s can be found 
and can be interpreted as either dark and cold plasma falling back to Betelgeuse, 
or to rising plumes of plasma located in the hidden face of Betelgeuse,
and rising so high that they appear above the limb.

@@

Figures \ref{LS intensity},\ref{LS Q},\ref{LS U} and \ref{LS linear polarization} show the Lomb-Scargle periodogram in normalised power of the LSD profile of respectively 
the intensity, stokes Q, stokes U and the linear polarization. 
%The abscissa is the frequency (in $\mathrm{days^{-1}}$), while the ordinate is the heliocentric radial velocity in km/s. 
For each figure, the upper panel is the Lomb-Scargle periodogram with data taken before the great dimming of Betelgeuse, that is from 2013 to the beginning of 2019. 
The lower panel is the Lomb-Scargle periodogram for the whole dataset, before and after the dimming: from 2013 to 2023. The white dashed lines mark the different 
periods found in the literature: the LSP at 2000 days, the fundamental pressure mode at 400 days and its first overtone mode at 200 days.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle Intensity.png}
    \caption{Lomb-Scargle periodogram of the LSD profile of intensity. 
    The upper panel is the Lomb-Scargle periodogram for data before the great dimming while the lower panel is the periodogram from the whole dataset.
    The ordinate refers to the velocity in the heliocentric radial velocity in km/s.
    The three white dashed lines represent respectively (from left to right) the LSP at 2000 days, the 400 days period and its first overtone at 200 days.  }
    \label{LS intensity}
\end{figure}

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle linear polarization.png}
    \caption{Same as figure \ref{LS intensity} for linear polarization: $\sqrt{\mathrm{Q^2}+\mathrm{U^2}}$.}
    \label{LS linear polarization}
\end{figure}

@@

%\subsection{Intensity}
Figure \ref{LS intensity} shows the Lomb-Scargle periodogram of the LSD of the intensity profile. Interestingly, we recover the LSP at 2000 days, 
but located at a HRV of 20 km/s rather than over the whole profile. This periodicity is still present after the  dimming, but  it is located at a HRV between 0 and 10 km/s. 
There is also a weaker signal at the fundamental mode, around 400 days before the dimming. But this signal completely disappears after the dimming. From these observations
we find  hard to attribute 
the observed Lomb-Scargle power to a global period of Betelgeuse.
%, however it might be a hint that this periodicity has changed after the dimming. 
Our results 
in the intensity profile are consistent with the one from \cite{mathias_evolution_2018} who used 2D Fourier analysis and found that the dominant period was the LSP.

@@

Figure \ref{LS Q} and \ref{LS U} show the Lomb-Scargle periodogram of both stokes Q and U. It is clear that the periodogram is completely different 
from the intensity one. A strong signal is present in fig. \ref{LS Q} around 200 days. Once again, this signal is weaker after the dimming.
Concerning stokes U in fig. \ref{LS U}, it appears that we only recover the LSP, which once again, is weaker after the great dimming. 
Those periodograms rise a lot of questions: why is the 200 days period only present in stokes Q? Why not in stokes U or intensity? 
We will discuss about those questions in the next section. It also appears that stokes Q is
much more sensitive to the 200 days period than the LSP. From this, we can conclude that stokes
Q evolves on a time scale much more closer to the first overtone 
mode than the LSP or even the fundamental mode. Since Stokes Q is sensitive to different periods, we must conclude that it is due to other mechanisms
than the one behind the periodicities observed in intensity. Linear polarization in Betelgeuse is interpreted to be a proxy of  convective cells,
so we can speculate that the signal present in the periodogram 
may be related to  the lifetime of convective cells. Such hypothesis is comforted by what we see in the 
LSD profiles of  Betelgeuse, whose shape is overhauled  in typical time scales of  a few months.
Such time scales observed in the Stokes profiles and attributed to the convective time scales 
appear to explain the strong signal present at 200 days in the Lomb-Scargle periodogram of Stokes Q.

@@

Figure \ref{LS linear polarization} shows the Lomb-Scargle periodogram of the total linear polarization of Betelgeuse: $\sqrt{\mathrm{Q^2+U^2}}$. 
Interestingly, the periodogram shows a strong power at 400 days and 2000 days. This signal is much more weaker after the dimming.
While the LSP is already present in fig. \ref{LS intensity}, here we recover the 400 days periodicity, often invoked in the literature. 
Nonetheless, the power of this period peaks at a HRV of around 40 km/s. We recall that this velocity is the rest velocity of the star. 
Every signal beyond this velocity is either due to  plasma sinking in Betelgeuse, or to plumes of plasma rising beyond and above the limb of the star. 
Such plumes from the back hemisphere have been often seen in other red supergiants like $\mu$ Cep \citep{lopez_ariste_height_2023}. 
This suggests that both signals at these period of 400 days found in intensity and total linear polarisation 
could well be due to the periodic appearance of plumes in the back hemisphere rising high enough to appear above
the visible limb and generating a supplementary source of brightness. Once again we find a plausible explanation 
of these periods unrelated to pulsations, but rather to the convective dynamics of the star.

@@

Using linear polarization, \cite{lopez_ariste_convective_2018} have been able to reconstruct images of Betelgeuse that have been successfully compared 
to inteferometric images of \cite{montarges_close_2016}. The images are produced by finding the brightness distribution that better fits the observed linear 
polarization LSD profile using a Marquardt-Levenberg minimisation.

Betelgeuse is observed in average every two or three weeks by the TBL, and hence
we are able to follow its surface activity through polarimetry imaging. This has  allowed in the past the estimation of  the size and the
 lifetime of the convective cells
at the surface. From these images, we computed a photo-center, which is sensitive to the size and the number of convective cells. 
An homogeneous star will have a photo-center displacement coinciding with barycenter of the star, whereas a star with one or two huge convective cells will have a
photo-center displacement more important, up to a few percent of the stellar radius in the case of red supergiants \citep{chiavassa_probing_2022}.

@@

%We computed the Lomb-Scargle periodogram of the displacement of the photo-center.
But, before going any further, we have to warn
 about the interpretation of such images. 
Linear polarization suffers from a $180 ^\circ$ degree ambiguity already mentioned in \cite{auriere_discovery_2016}. For that reason, our images 
can be 
rotated by $180^\circ$ degree, and the brightness distribution will still fit the observed LSD polarization profiles. 
If a particular observation were to be treated independently of previous observations, the solution found by 
the algorithm could be any of the ambiguous solutions possible. To obtain a continuity 
between the series of images, 
for a given day we used the brightness distribution of the previous day as initial point 
of the  fitting iteration; and for the first image, we start with a 
random brightness distribution as the beginning of the fit. 
Even though we are able to produce a series of images  consistent with each other, it is important 
to keep in mind that the photo-center displacement computed from the series 
will be affected by the choice of the first image. 
To overcome this issue, we decided to compute
the photo-center displacement from 100 different initial parameters and their subsequent series 
of images in an effort to recover from the ensemble
properties that are independent of the choice of the first image.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle Photo-center.png}
    \caption{Lomb-Scargle periodogram of the 100 photo-center displacement of Betelgeuse. Each blue curve corresponds to a Lomb-Scargle periodogram of one photo-center
     displacement. The black line is the average of the 100 periodogram. The black dashed lines represent the 2000, 400 and 200 days period respectively 
     (from left to right). In the upper panel, only data before the dimming were used to compute the Lomb-Scargle periodogram, while for the lower panel, the whole dataset was used. }
    \label{LS photocenter}
\end{figure}

Figure \ref{LS photocenter} shows the Lomb-Scargle periodogram of the photo-center displacement of each one of those 100 series (blue lines) and the average Lomb-Scargle periodogram (black line). 
As before, the upper panel shows the Lomb-Scargle periodogram from data before the great dimming only, whereas we used the whole set of data for the lower panel.
The black dashed lines mark the 2000, 400 and 
200 days periods respectively. First, we find that the Lomb-Scargle power is high around the LSP, though
uncomfortably close to the auto-correlation peak.
We also notice a decrease in the Lomb-Scargle power from  before to after the dimming, in 
concurrence to what was seen in the periodograms of the LSD profiles. Before it, there are two smalls bumps around the 400 days period. 
But those bumps almost completely disappear after the dimming. Even though it is 
hard to identify the 400 days period from those bumps alone,
we still find that there is a difference before and after the dimming. 
And this is concluded  from images that exclusively involve surface convection.
The Lomb-Scargle periodogram of the photo-center displacement points towards a variability explained by convection only,
and this is consistent with what we obtained from linear polarization.
%Both the photo-center displacement and the LS of the linear polarization point toward a variability explained by surface convection.

@@

We have recovered the light curve of Betelgeuse in the visible from the AAVSO database for the last 13 years (Figure \ref{light curve Betelgeuse}).
Before the great dimming, the magnitude of Betelgeuse evolves on a timescale of a year, whereas after the dimming, the variability of Betelgeuse has changed. 
It is clear from fig. \ref{light curve Betelgeuse} that the 
variability of Betelgeuse is evolving on a timescale shorter than one year. This qualitative change has been pointed out before.
In the literature, it is reported that before the great dimming the main period of Betelgeuse was of 400 days \citep{kiss_variability_2006}. 
This period is often associated to the fundamental pressure mode. After the dimming, however this period appears to have disappeared 
and only the period at 200 days,  associated to the first overtone of the fundamental mode, is visible since \citep{dupree_great_2022}.
No explanations based upon pulsations have been offered concerning this change of variability after or perhaps because of the dimming. But this 
is a behaviour that fully concurs with what we find in the polarization data, and which cannot be related to any stellar pulsation but rather 
convective dynamics. Beyond the actual values of the periods found in a Lomb-Scargle periodogram, we also see a correspondence between polarization 
and the light curve in these changes in the observed periods before and after the great dimming. Such concurrence points to an explanation 
in terms of convective dynamics also for the variability of the light curve.

It may be argued that the periodograms presented in the previous sections and, in particular, those  of the photo-center displacements, 
present low amplitude peaks at the referred main periods. One may wonder whether we are over-interpreting mere fluctuations of the periodograms, 
while the peaks seen in the literature from analysis of the light curve are unmistakable. It is therefore worth to take advantage of the 
availability of the AAVSO data to perform a Lomb-Scargle periodogram of the light curve in the same period covered by our polarization data and,
furthermore, taking into consideration data around the dates for which TBL data is available. Periodograms  with such constraints are 
shown in Fig.\ref{LSAAVSO}. Even when the full dataset for the period is taken into account (red line), the periods present small amplitudes, and are 
fully and favourably comparable with the periodograms presented above. This is even more dramatic when the periodogram is made from data at the exclusive 
dates at which TBL data is available (blue line). It stands out that the low amplitude peaks are rather due to both the sparse cadence and short span of 
time of the TBL data series rather than to a physical absence. Actually, the present tests puts more emphasis on the importance of the 
strong peaks seen at selected wavelengths in the LSD profiles, strong when compared to the peaks at similar periods arising from the light curve.

\begin{figure}[!h]
    \centering
    \includegraphics[width=\linewidth]{Lomb-Scargle-TBL.png}
    \caption{Lomb-Scargle periodogram from the AAVSO data between 2013 and 2023 (red line) and from the AAVSO data taken near TBL observations (blue line).}
    \label{LSAAVSO}
\end{figure}

@@

The interpretation of the linear polarization profiles is made in terms of convective structures in the atmosphere of Betelgeuse. We suggest 
that it is in these convective structures and their temporal evolution that we will find the true reason for the observed periodicities both 
in the light curve and in the spectropolarimetric observations. This is the main conclusion of this work: the observed periods are related 
to convective dynamics and not to radial pulsations of the star.
